Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-06-08 analysis/paper/paper.Rmd
Corrections after seminar

Even though each individual assemblage can have been impacted by an virtual infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the so-called whole assemblage behavioural index [@clark2017], which has not seen application in this context before.

@@

The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway \@ref(fig:dist)A. The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b]. The sites included in the analysis consists of all Mesolithic sites excavated in conjunction with the projects that have assemblages holding more than 100 artefacts. The institution responsible for these excavations was the Museum of Cultural History in Oslo. This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations. Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the museum [@melvold2014c; @koxvold2017]. As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [e.g. @clark2006; @dibble2017]. In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014, see also \@ref(fig:dist)B]. Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages. While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects.

```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis. Radiocarbon age determintations are given as the sum of the posterior density esimates. Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators. These follow the original reports.", fig.height = 13, fig.width = 12}
# Create a bounding box around site points to limit the overview map
sitbbox <- st_bbox(site_pts)
sitbbox[1:2] <- sitbbox[1:2] - 600000
sitbbox[3:4] <- sitbbox[3:4] + 600000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [e.g. @bjerck2008]. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [e.g. @åstveit2018; @breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of relative sea-level change allows for a relatively good chronological control of when these accumulation events occurred. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year towards the end of the Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.

@@

However, following not least from the ambiguous definition first put forward by @binford1973, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed and disputed, and no single definition has ever been reached [e.g. @bamforth1986; @nash1996; @shott1996; @surovell2009, 9--13]. The continuum between curated and expedient technology has for example been related to dimensions such as land-use and mobility strategies and raw-material quality and availability [e.g. @parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017]. Still, that the distinction can offer a useful analytical point of departure if clearly and explicitly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted. For example, although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool --- or extending its use-life --- by the repeated rejuvenation and modification of edges [e.g. @bamforth1986; @dibble1995; @shott2005]. Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g. @barton1999; @riel-salvatore2004; @barton2011; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @kuhn2015]. This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014]. In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to an expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades. These variables and underlying logic consitute what has been termed the whole assemblage behavioural index [WABI, @clark2017], and is the main framework adopted here. However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g. @breivik2016a; @reitan2016; @Bicho2020; @kitchel2021]. The use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181], while chips . In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material. The analysis is done using principal components analysis (PCA), following the continuous measure of these variables [cf. @baxter1994, 100; @Bicho2020].

@@

Figure \@ref(fig:cor) displays the CA using the lithic count data. The general impression from the plots is that a chronological dimension is associated with the patterning in the data. This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, and the horseshoe curve or Guttman effect evident in the column plot [see @baxter1994, 119--120]. The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension. Although fewer in number, the sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to a larger extent are characterised by core fragments, both in flint and non-flint materials, rather than the cores themselves. The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites. Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014]. However, as the site had been quite heavily impacted by modern disturbances, this led [@eigeland2014, 124] to suggest that the material might have been compromised. This could explain its position as an outlier in the plot. The use of metarhyolite and smoky quartz for the production of axes is present at other contemporary sites as well, but is evidently not as prominent a part of these assemblages [@jaksland2014b]. In sum, the findings for the earliest sites are in large part in line with previous research [e.g. @jaksland2014b; @breivik2018; @damlien2018b].

@@

Moving on the PCA of measures that have been linked to mobility, some of the variables with severely skewed distributions were initially transformed. These are displayed in the correlation matrix in in Figure \@ref(fig:cor-plot). Figure \@ref(fig:pca-analysis) displays the resulting PCA. There is a general temporal transition from older to younger sites from the upper left to the bottom right of the plot. The second dimension is mainly defined by a negative correlation between the VDL and RFSL (Figure \@ref(fig:var-contrib)). Almost orthogonal to this is the strong negative correlation between relative frequency of chips and blanks. While there is a slight tendency for blanks to be more associated with younger sites, frequency of chips appears to be largely independent of time. However, the almost suspiciously strong negative correlation between chips and blanks can perhaps have a practical explanation. Seeing as the frequency of non-flint material is positively correlated with blanks and negatively correlated with chips (Figure \@ref(fig:cor-plot), one explanation to this pattern could be that smaller non-flint pieces are simply more difficult to identify and separate from naturally fragmented stone during excavation and classification. This could conceivably have led to an over-representation of blanks as compared to chips in assemblages with a high proportion of non-flint material. While this is not necessarily the entire explanation behind the relationship, this does make it difficult to place much analytical weight on this pattern. Relative frequency of cores is not especially impactful in the PCA, and appears to be independent the temporal dimension as well. That is not to say that cores may not be indicative or related to mobility patterns, but to get at this may require further analysis beyond their simple classification as cores [@kitchel2021].

@@

Figure \@ref(fig:wabi)A illustrates the negative correlation between the two variables (*r =* `r round(wabicor, 2)`) while also displaying a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites. The linear correlation is stronger between the mean site age and RFSL (*r =* `r round(secfreqcor, 2)`), than between mean site age and VDL (*r =* `r round(denscor, 2)`). As variable non-flint availability and workability has also been suggested to potentially impact these measures [@manninen2014], Figure \@ref(fig:wabi)B displays the same relationship, but exclusively for the flint data. While the negative correlation is slightly less marked when only the flint data is considered (*r =* `r round(wabicorflint, 2)`), the general pattern is the same. The relationship between mean site age and relative frequency of secondarily worked flint (RFSF) is even stronger (*r =* `r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the x-axis, the volumetric density of flint (VDF) is not temporally contingent (*r =* `r round(denscorflint, 2)`). As was also indicated by the CA, this follows from the fact that non-flint materials make up a higher share of the assemblages for some of the later Mesolithic sites, and is a point returned to below where the temporal dimension of the relationship between VDL and RFSL is explored further.

@@

The results of the CA does appear to align well with previous research, and the employed artefact categories are clearly capturing a temporal component. One possible implication of this could be that the aggregation of artefact categories might have been overly conservative. However, it also evidently clear, in the words of @kruskal1971 [22], that 'time is not the only dimension'. The results of the CA do most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, which is also made evident by some of significant deviances from the overall pattern. Unpicking and aligning these patterns with any specific behavioural and technological dimensions using the coarse CA results is, however, another task entirely. This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural and social significance of the employed units of analysis unclear. The results of the CA can, however, be used in conjunction with the part of the analysis that has attempted to get at more specific behavioural dimensions to nuance or explain discrepancies in this data.

@@

While there are certainly nuances and idiosyncrasies in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust. It seems clear that there is a marked drop starting some time just before 8000 BCE, which has stabilised around 7000 BCE. This corresponds well with a chronological framework where the end of the Early Mesolithic, or Flake axe phase, is set to c. 8200 BCE [@reitan2016]. Previous research has proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @fuglestvedt2012]. This corresponds very well with the findings reported here, where the earliest assemblages are characterised by relatively high and uniform values on the curation index. The transition to the subsequent Middle Mesolithic, or Microlith phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology, which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian Steppes [@günther2018; @manninen2021]. The radiocarbon record points towards a coinciding population decline in Southern Norway around this time [@nielsen2021]. Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021]. In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE. Referring back to the increasing expediency in the curation data between c. 8200 and 7000 BCE, the Microlith phase could thus represent a transitional phase where migrating people and new living practices was propagating through societies in South-Eastern Norway --- a process that in light of the curation data would have concluded around 7000 BCE.

@@

The subsequent Transverse arrowhead phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged -and single-edged points [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021]. The continued stability of the curation index could indicate that these changes are not related to major shifts in land-use and mobility patterns related to the sites. However, it is also worth highlighting the strained sample size for the later parts of the Mesolithic here, which could mean that the effect could be missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.

As it stands, however, the main hypotheses resulting from the analysis would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high degrees of mobility, which then drop before leveling off at around 7000 BCE. These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA. Although the precise nature of this transition would require further consideration, the quite dramatic fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a residential to a logistical settlement system [@barton2011; @binford1980].

@@

The results of the CA using the entirety of the lithic assemblages align more or less with results of previous research in south-eastern Norway. This would indicate that in general, meaningful chronological patterning is associated with the employed artefact categories. These tendencies are already well-established when it comes to the formal tool types, but have been given less focus in light of entire assemblages. Precisely what behavioural implication the development in the use of these types have, however, is unclear, but appears to follow a different and more complex trajectory than that of curation, as measured here.

>>>>

Even though each individual assemblage can have been impacted by an virtual infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the so-called whole assemblage behavioural index [@clark2017], which aligns properties of the assemblages with has not seen application in this context before.

@@

The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway (Figure \@ref(fig:dist)A). The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b]. The sites included in the analysis consist of all Mesolithic sites excavated in conjunction with the projects that have assemblages holding more than 100 artefacts. The institution responsible for these excavations was the Museum of Cultural History in Oslo. This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations. Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the museum [@melvold2014c; @koxvold2017]. As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [e.g. @clark2006; @dibble2017]. In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@jaksland2014; @solheim2013b, see also Figure \@ref(fig:dist)B]. Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages. While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects.

```{r dist, fig.cap = "A) Spatial and B) temporal distribution of the sites chosen for analysis. Radiocarbon age determinations are given as the sum of the posterior density esimates. Solid lines indicate that the site has been dated with reference to relative sea-level change and typological indicators. These follow the original reports.", fig.height = 13, fig.width = 12}
# Create a bounding box around site points to limit the overview map
sitbbox <- st_bbox(site_pts)
sitbbox[1:2] <- sitbbox[1:2] - 600000
sitbbox[3:4] <- sitbbox[3:4] + 600000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [e.g. @bjerck2008]. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [e.g. @åstveit2018; @breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of relative sea-level change allows for a relatively good control of when these accumulation events occurred. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of losing temporal resolution as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year towards the end of the Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared --- a point that is returned to in the discussion of the results.

@@

However, following not least from the ambiguous definition first put forward by @binford1973, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed and disputed, and no single definition has ever been reached [e.g. @bamforth1986; @nash1996; @shott1996; @surovell2009, 9--13]. The continuum between curated and expedient technology has for example been related to dimensions such as land-use and mobility strategies and raw-material quality and availability [e.g. @parry1987; @kuhn1992; @andrefsky1994; @smith2015; @clark2017]. Still, that the distinction can offer a useful analytical point of departure if clearly and explicitly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted. For example, although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool --- or extending its use-life --- by the repeated rejuvenation and modification of edges [e.g. @bamforth1986; @dibble1995; @shott2005].

One concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g. @barton1999; @riel-salvatore2004; @barton2011; @barton2014; @clark2017], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @kuhn2015]. This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014]. In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to a more expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades. These variables and underlying logic consitute what has been termed the whole assemblage behavioural index [WABI, @clark2017], and is the main framework adopted here. However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g. @breivik2016a; @reitan2016; @Bicho2020; @kitchel2021]. The use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181], while chips . In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material. Following [@Bicho2020] the analysis is done using principal components analysis (PCA), leading to a [cf. @baxter1994, 100].

@@

Figure \@ref(fig:cor) displays the CA using the lithic count data. The general impression from the plots is that a chronological dimension is associated with the patterning in the data. This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), combined with the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [see @baxter1994, 119--120]. The earliest sites tend to be located in the upper right corner of plot A, with increasingly younger sites towards the bottom along the second dimension. Although fewer in number, the sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to a larger extent are characterised by core fragments, both in flint and non-flint materials, rather than the cores themselves. The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites. Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014]. However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014, 124 to suggest that the material might have been compromised. This could explain its position as an outlier in the plot. The use of metarhyolite and smoky quartz for the production of axes is present at other contemporary sites as well, but is evidently not as prominent a part of these assemblages [@jaksland2014b]. In sum, the findings for the earliest sites are in large part in line with previous research [e.g. @jaksland2014b; @breivik2018; @damlien2018b].

@@

Moving on to the PCA of measures that have been linked to mobility, some of the variables with severely skewed distributions were initially transformed. These are displayed in the correlation matrix in in Figure \@ref(fig:cor-plot). Figure \@ref(fig:pca-analysis) displays the resulting PCA. There is a general temporal transition from older to younger sites from the upper left to the bottom right of the plot. The second dimension is mainly defined by a negative correlation between the VDL and RFSL (Figure \@ref(fig:var-contrib)). Almost orthogonal to this is the strong negative correlation between relative frequency of chips and blanks. While there is a slight tendency for blanks to be more associated with younger sites, frequency of chips appears to be largely independent of time. However, the almost suspiciously strong negative correlation between chips and blanks can perhaps have a practical explanation. Seeing as the frequency of non-flint material is positively correlated with blanks and negatively correlated with chips (Figure \@ref(fig:cor-plot), one explanation to this pattern could be that smaller non-flint pieces are simply more difficult to identify and separate from naturally fragmented stone during excavation and classification. This could conceivably have led to an over-representation of blanks as compared to chips in assemblages with a high proportion of non-flint material. While this is not necessarily the entire explanation behind the relationship, this does make it difficult to place much analytical weight on this pattern. Relative frequency of cores is not especially impactful in the PCA, and appears to be independent the temporal dimension as well. That is not to say that cores may not be indicative or related to mobility patterns, but to get at this may require further analysis beyond their simple classification as cores [@kitchel2021].

@@

Figure \@ref(fig:wabi)A illustrates the negative correlation between the two variables (*r =* `r round(wabicor, 2)`) while also displaying a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites. The linear correlation is stronger between the mean site age and RFSL (*r =* `r round(secfreqcor, 2)`), than between mean site age and VDL (*r =* `r round(denscor, 2)`). As variable non-flint availability and workability has also been suggested to potentially impact these measures [@manninen2014], Figure \@ref(fig:wabi)B displays the same relationship, but exclusively for the flint data. While the negative correlation is slightly less marked when only the flint data is considered (*r =* `r round(wabicorflint, 2)`), the general pattern is the same. The relationship between mean site age and relative frequency of secondarily worked flint is even stronger (*r =* `r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the x-axis, the volumetric density of flint is not temporally contingent (*r =* `r round(denscorflint, 2)`). As was also indicated by the CA, this follows from the fact that non-flint materials make up a higher share of the assemblages for some of the later Mesolithic sites, and is a point returned to below where the temporal dimension of the relationship between VDL and RFSL is explored further.

@@

The results of the CA does appear to align well with previous research, and the employed artefact categories are clearly capturing a temporal component. One possible implication of this could be that the aggregation of artefact categories might have been overly conservative. However, it is also evidently clear, in the words of @kruskal1971 [22], that 'time is not the only dimension'. The results of the CA do most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, which is also made evident by some of significant deviances from the overall pattern. Unpicking and aligning these patterns with any specific behavioural and technological dimensions using the coarse CA results is, however, another task entirely. This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural and social significance of the employed units of analysis unclear. The results of the CA can, however, be used in conjunction with the part of the analysis that has attempted to get at more specific behavioural dimensions to nuance or explain discrepancies in this data.

@@

While there are certainly nuances in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust. It seems clear that there is a marked drop starting some time just before 8000 BCE, which has stabilised around 7000 BCE. This corresponds well with a chronological framework where the end of the Early Mesolithic, or Flake axe phase, is set to c. 8200 BCE [@reitan2016]. Previous research has proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @fuglestvedt2012]. This corresponds very well with the findings reported here, where the earliest assemblages are characterised by relatively high and uniform values on the curation index. The transition to the subsequent Middle Mesolithic, or Microlith phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology, which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian Steppes [@günther2018; @manninen2021]. The radiocarbon record points towards a coinciding population decline in Southern Norway around this time [@nielsen2021]. Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021]. In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE. Referring back to the increasing expediency in the curation data between c. 8200 and 7000 BCE, the Microlith phase could thus represent a transitional phase where migrating people and new living practices were propagating through societies in South-Eastern Norway --- a process that in light of the curation data would have concluded around 7000 BCE.

@@

The subsequent Transverse arrowhead phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged -and single-edged points [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021]. The continued stability of the curation index could indicate that these changes are not related to major shifts in land-use and mobility patterns in the material treated here. However, it is also worth highlighting the strained sample size for the later parts of the Mesolithic, which could mean that the effect is missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.

As it stands, however, the main hypotheses resulting from the analysis would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high and unform degrees of mobility, which then drop before leveling off at around 7000 BCE. These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA. Although the precise nature of this transition would require further consideration, the quite dramatic fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a predominantly residential to logistical settlement system [@barton2011; @binford1980].

@@

The results of the CA using the entirety of the lithic assemblages align more or less with results of previous research in south-eastern Norway. This would indicate that in general, meaningful chronological patterning is associated with the employed artefact categories. These tendencies are already well-established when it comes to the formal tool types, but have been given less focus in light of entire assemblages. Precisely what behavioural implication the development in the use of these types have, however, is unclear, but appears to follow a different and more complex trajectory than that of curation, as operationalised here.
